\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-12 paper/ms.tex
snark about model selection

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost commonly scales as the cube of the number of data points,
their application has been limited to relatively small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
A point estimate for the values of the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
\response{Furthermore}, probabilistic constraints on $\bvec{\theta}$ and
$\bvec{\alpha}$ can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets, $N \approx 10^3$, because the cost of
computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
$\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
This means that for large datasets, every evaluation of the likelihood quickly
becomes computationally intractable, and representing the matrix in memory can
become expensive.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celeriteterm\ terms.

@@

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are restricted to one-dimensional problems and the likelihood
function for a CARMA model can be solved in $\mathcal{O}(N\,J^2)$ using a
Kalman filter \citep{Kelly:2014}.
The kernel function for a CARMA$(J,\,K)$ model is
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model; $\sigma$ is real, while the others are complex.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
however, in practice this method is somewhat slower.
An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.

\begin{figure}[tp]
\begin{center}
\includegraphics[width=\textwidth]{figures/benchmark_darwin_carma.pdf}
\caption{The same as \Figure{benchmark}, but using an optimized
    \project{C++} implementation of the Kalman filter method developed by
    \citet{Kelly:2014}.
    Comparing this figure to \Figure{benchmark}, we see that \celerite\ is
    about an order of magnitude faster in all cases.
    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
    matrix given by \eq{celerite-kernel} as a function of the number of data
    points $N$.
    The different lines show the cost for different numbers of terms $J$
    increasing from bottom to top.
    To guide the eye, the straight black line without points
    shows linear scaling in $N$.
    \emph{(right)} The same information plotted as a function of $J$ for
    different values of $N$.
    Each line shows the scaling for a specific value of $N$ increasing from
    bottom to top.
    The black line shows quadratic scaling in $J$.
    \figurelabel{benchmark-carma}}
\end{center}
\end{figure}

\subsection{Other scalable methods}

@@

Another GP method that has been used extensively in astronomy is the
hierarchical off-diagonal low rank \citep[HODLR,][]{Ambikasaran:2016} solver.
This method exploits the fact that many commonly used kernel functions produce
``smooth'' matrices to approximately compute the GP likelihood with the
scaling $\mathcal{O}(N\,\log^2 N)$.
This method has the advantage that, unlike \celerite, it can be used with any
kernel function but, in practice, the cost can still prove to be prohibitively
high for multi-dimensional inputs.
The proportionality constant in the $N\,\log^2N$ scaling of the HODLR method
is a function of the specific kernel and we find~--~using the \project{george}
software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
scales approximately linearly with $J$, but it requires substantial overhead
for small models.
This means that the HODLR solver can \emph{approximately} evaluate
\celeriteterm\
models more efficiently than \celerite\ for large models $J \gg 10$ and
small datasets where $N$ is smaller than a few thousand.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost generally
scales as the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
oscillator matches simple models of astrophysical variability, and makes the
parameterization interpretable in terms of resonant frequency, amplitude, and
quality factor.
% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in some cases, small values of $J$ may suffice.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic data analysis that are computationally tractable, even with
large datasets.
Gaussian Processes are a popular class of models used for this purpose but,
since the computational cost scales, in general, as the cube of the number of
data points, their application has been limited to small datasets.
In this paper, we present a novel method for Gaussian Process modeling in
one-dimension where the computational requirements scale linearly with the
size of the dataset.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it can be a useful effective model in some other cases.
We present a mathematical description of the method and compare it to existing
scalable Gaussian Process methods.
The method is fast and interpretable, with a wide range of potential
applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
\response{
A point estimate for the values of the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.}
\response{Furthermore, the uncertainties on $\bvec{\theta}$ and
$\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.}

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets, \response{$N \lesssim
10^3$}, because the cost of computing the inverse and determinant of a general
matrix $K_\bvec{\alpha}$ is $\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
This means that for large datasets, every evaluation of the likelihood quickly
becomes computationally intractable, and representing the matrix in memory can
become expensive.
In this case, the use of non-linear optimization or MCMC is no longer
practical.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in \sect{factor} can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celeriteterm\ terms.

@@

\subsection{How to choose a celerite model}

In any application of \celerite~--~or any other GP model, for that
matter~--~one must choose a kernel.
A detailed discussion of how to choose a kernel is beyond the scope of this
paper and the interested reader is directed to Chapter~5 of
\citet{Rasmussen:2006} a general introduction, but we include a summary and
some suggestions specific to \celerite\ here.
We argue that the most satisfying method of choosing a kernel is to use
physical motivation and the discussion in \sect{sho} provides an opportunity
for justified model specification when \celerite\ is applied to parameter
estimation from stellar light curves.
However, it is not always possible to rely on physics to specify the model
and, in these cases, it is useful to have other tools at one's disposal.
One option is to compute the evidence integral or Bayes factor for a set of
models and choose the model with the highest evidence.
Another option is to perform cross validation and select the model with the
best predictive performance.
For continuous autoregressive moving average models (we return to these in
\sect{compare}), \citet{Kelly:2014} recommend the Bayesian Information
Criterion (BIC) as a method for selecting the order of the model.
This is a total hack, but I guess that you could do it if you wanted.

@@

The continuous autoregressive moving average (CARMA) models introduced to the
astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are restricted to one-dimensional problems and the likelihood
function for a CARMA model can be solved in $\mathcal{O}(N\,J^2)$ using a
Kalman filter \citep{Kelly:2014}.
The kernel function for a CARMA$(J,\,K)$ model is
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model; $\sigma$ is real, while the others are complex.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.

\response{
The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
the Kalman filter method \citep{Kelly:2014}, but we find that, in practice,
this method is more computationally expensive than \celerite, but it has a
smaller memory footprint.
\Figure{benchmark-carma} shows the scaling of the Kalman filter solver for the
systems shown in \Figure{benchmark}.
Comparing these figures, we find that the Kalman filter solver is slower than
\celerite\ for all the systems we tested with an average difference of about
an order of magnitude.

Another benefit of \celerite\ compared to Kalman filter solvers is the fact
that it is a direct solver for the relevant matrix equations instead of just
an algorithm to compute \eq{gp-likelihood}.
This means that the inverse covariance matrix $K^{-1}$ can be applied to
general vectors and matrices in $\mathcal{O}(N)$, while reusing the
factorization.
This is a crucial feature when GP models are combined with linear regression
\citep[for example][]{Luger:2017}.
}

\begin{figure}[tp]
\begin{center}
\includegraphics[width=\textwidth]{figures/benchmark_darwin_carma.pdf}
\caption{\response{The same as \Figure{benchmark}, but using an optimized
    \project{C++} implementation of the Kalman filter method developed by
    \citet{Kelly:2014}.
    Comparing this figure to \Figure{benchmark}, we see that \celerite\ is
    about an order of magnitude faster in all cases.
    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
    matrix given by \eq{celerite-kernel} as a function of the number of data
    points $N$.
    The different lines show the cost for different numbers of terms $J$
    increasing from bottom to top.
    To guide the eye, the straight black line without points
    shows linear scaling in $N$.
    \emph{(right)} The same information plotted as a function of $J$ for
    different values of $N$.
    Each line shows the scaling for a specific value of $N$ increasing from
    bottom to top.
    The black line shows quadratic scaling in $J$.}
    \figurelabel{benchmark-carma}}
\end{center}
\end{figure}

@@

Another GP method that has been used extensively in astronomy is the
hierarchical off-diagonal low rank (HODLR) solver \citep{Ambikasaran:2016}.
This method exploits the fact that many commonly used kernel functions produce
``smooth'' matrices to approximately compute the GP likelihood with the
scaling $\mathcal{O}(N\,\log^2 N)$.
This method has the advantage that, unlike \celerite, it can be used with any
kernel function but, in practice, the cost can still prove to be prohibitively
high for multi-dimensional inputs.
The proportionality constant in the $N\,\log^2N$ scaling of the HODLR method
is a function of the specific kernel and we find~--~using the \project{george}
software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
scales approximately linearly with $J$, but it requires substantial overhead
making it much slower than \celerite\ for all the models we tested.
For large $J \gtrsim 256$ and small $N \lesssim 1000$, we find that
\project{george} can approximately evaluate \celeriteterm\ models with the
same or less computational cost than \celerite, but that \celerite\ is
faster~--~and exact~--~in all other parts of parameter space.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost generally
scales as the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped,
stochastically-driven harmonic oscillator matches simple models of
astrophysical variability, and makes the parameterization interpretable in
terms of resonant frequency, amplitude, and quality factor.
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in some cases, small values of $J$ may suffice.
